---
title: "Integrated families project.v2.0.rmd"
output: html_notebook
---
```{r}
#This project requires 4 plugins, which can be installed with this code:
install.packages("odbc")
install.packages("DBI")
install.packages("openxlsx")
install.packages("tidyverse")
```

```{r}
#Loading libraries. After the packages above are installed, they need to be loaded. They only need to be installed on a computer once, but they need to be loaded every time RStudio is opened.
#The memory.limit function was required on the computer I used to increase the memory size so that large tables could be loaded. It may not be needed in all cases.
library(odbc)
library(DBI)
library(openxlsx)
library(tidyverse)
memory.limit(size=56000)
```

```{r}
#This code connects to the HistoTrac database. The computer RStudio is running on needs to have access to HistoTrac, and the settings for your particular HistoTrac configuration may be different.

# Connecting to the HistoTrac database
con <- dbConnect(odbc::odbc(), "Histotrac", timeout = 10)

#Pulling a list of available tables in HistoTrac and adding it to "HTtableList"
(HTtableList <- dbListTables(con) %>% as_tibble)

#This will show all of the tables in the HistoTrac database. Not all of them are used, and only a few are required to search for families.
```
#The following code chunks will extract each of the necessary tables from HistoTrac, load them into R in the computer's memory, and then write them to a csv file in the active directory.
```{r}
#Extract and save Sample table. Use as a template for extracting and saving tables; just change the table names in the second and third lines of code.
#This uses the direct extraction code. If this doesn't work, use the error-correcting code based on the "Patient" table.
con <- dbConnect(odbc::odbc(), "HistoTrac", timeout = 10)
(dtSample <- dbReadTable(con, "Sample") %>% as_tibble)
dtSample %>% write_csv("Sample.csv.gz")
```

```{r}
#save RelatedPatient table
con <- dbConnect(odbc::odbc(), "HistoTrac", timeout = 10)
(dtRelatedPatient <- dbReadTable(con, "RelatedPatient") %>% as_tibble)
dtRelatedPatient %>% write_csv("RelatedPatient.csv.gz")
```

```{r}
#Extract and save Patient table. The Patient table was difficult to get out, so error-correcting code was used.  
#This uses the error-correcting code for tables that do not extract easily. Only the last two lines have to be changed to extract and save different tables.
con <- dbConnect(odbc::odbc(), "HistoTrac", timeout = 10)
testpull <- function(con, tbnm, rownumber, ls_colnm, sp, ep) {
  re <- tryCatch({
    # try to query database based on given start point and end point of column
    dt <- dbGetQuery(con, paste("SELECT TOP", rownumber, paste0(ls_colnm[sp:ep], collapse = ', '), "FROM", tbnm))
    return(dt)
  },
  error = function(error_message) {
    return("error")
  },
  warning = function(warning_message) {
    return("warning")
  }
  )
}
pull <- function(con, tbnm) {
  # get total row number based on table name
  rownumber <- dbGetQuery(con, paste("SELECT count(*) FROM", tbnm))[[1]]
  # get all column names of the table
  ls_colnm <- c(dbGetQuery(con, paste("SELECT COLUMN_NAME FROM INFORMATION_SCHEMA.COLUMNS WHERE TABLE_NAME='",tbnm,"'",sep=""))$COLUMN_NAME)
  # set up variables for iteration
  colnumber <- length(ls_colnm)
  mincolnbr <- 1
  maxcolnbr <- colnumber
  sp <- mincolnbr
  ep <- maxcolnbr
  # initiate an empty dataframe
  df_total <- setNames(data.frame(matrix(nrow=rownumber, ncol=1)),c("x"))
  
  while (dim(df_total)[2] < maxcolnbr) {
    #print(c(sp,ep))
    df_test <- testpull(con, tbnm, rownumber, ls_colnm, sp, ep)
    #print(typeof(df_test))
    if (typeof(df_test) == "list") {
      # save successfully pulled data
      if (c(colnames(df_total))[1] == "x") {
        df_total <- subset(cbind(df_total, df_test),select=-c(x))
      }
      else {
        df_total <- cbind(df_total, df_test)
      }
      if (dim(df_total)[2] < maxcolnbr) {
        # adjust both start and end points when testpull succeeded
        sp <- ep + 1
        ep <- maxcolnbr
        #print(c(sp, ep))
      }
    }
    else {
      dbDisconnect(con)
      con <- dbConnect(odbc::odbc(), "HistoTrac", timeout = 10)
      # adjust end point when testpull failed
      ep <- floor(sp-1+((ep-(sp-1))/2))
      #print(c(sp, ep))
    }
  }
  return(df_total)
}

(dtPatient <- pull(con, "Patient") %>% as_tibble)
dtPatient %>% write_csv("Patient.csv.gz")
```

```{r}
#Save PatientCase table
con <- dbConnect(odbc::odbc(), "HistoTrac", timeout = 10)
testpull <- function(con, tbnm, rownumber, ls_colnm, sp, ep) {
  re <- tryCatch({
    # try to query database based on given start point and end point of column
    dt <- dbGetQuery(con, paste("SELECT TOP", rownumber, paste0(ls_colnm[sp:ep], collapse = ', '), "FROM", tbnm))
    return(dt)
  },
  error = function(error_message) {
    return("error")
  },
  warning = function(warning_message) {
    return("warning")
  }
  )
}
pull <- function(con, tbnm) {
  # get total row number based on table name
  rownumber <- dbGetQuery(con, paste("SELECT count(*) FROM", tbnm))[[1]]
  # get all column names of the table
  ls_colnm <- c(dbGetQuery(con, paste("SELECT COLUMN_NAME FROM INFORMATION_SCHEMA.COLUMNS WHERE TABLE_NAME='",tbnm,"'",sep=""))$COLUMN_NAME)
  # set up variables for iteration
  colnumber <- length(ls_colnm)
  mincolnbr <- 1
  maxcolnbr <- colnumber
  sp <- mincolnbr
  ep <- maxcolnbr
  # initiate an empty dataframe
  df_total <- setNames(data.frame(matrix(nrow=rownumber, ncol=1)),c("x"))
  
  while (dim(df_total)[2] < maxcolnbr) {
    #print(c(sp,ep))
    df_test <- testpull(con, tbnm, rownumber, ls_colnm, sp, ep)
    #print(typeof(df_test))
    if (typeof(df_test) == "list") {
      # save successfully pulled data
      if (c(colnames(df_total))[1] == "x") {
        df_total <- subset(cbind(df_total, df_test),select=-c(x))
      }
      else {
        df_total <- cbind(df_total, df_test)
      }
      if (dim(df_total)[2] < maxcolnbr) {
        # adjust both start and end points when testpull succeeded
        sp <- ep + 1
        ep <- maxcolnbr
        #print(c(sp, ep))
      }
    }
    else {
      dbDisconnect(con)
      con <- dbConnect(odbc::odbc(), "HistoTrac", timeout = 10)
      # adjust end point when testpull failed
      ep <- floor(sp-1+((ep-(sp-1))/2))
      #print(c(sp, ep))
    }
  }
  return(df_total)
}

(dtPatientCase <- pull(con, "PatientCase") %>% as_tibble)
dtPatientCase %>% write_csv("PatientCase.csv.gz")
```

```{r}
#Extract Test table. The Test table had similar problems, and so the error-correcting code was used. This is one of the largest tables, and may take a while to extract.
con <- dbConnect(odbc::odbc(), "HistoTrac", timeout = 10)
testpull <- function(con, tbnm, rownumber, ls_colnm, sp, ep) {
  re <- tryCatch({
    # try to query database based on given start point and end point of column
    dt <- dbGetQuery(con, paste("SELECT TOP", rownumber, paste0(ls_colnm[sp:ep], collapse = ', '), "FROM", tbnm))
    return(dt)
  },
  error = function(error_message) {
    return("error")
  },
  warning = function(warning_message) {
    return("warning")
  }
  )
}
pull <- function(con, tbnm) {
  # get total row number based on table name
  rownumber <- dbGetQuery(con, paste("SELECT count(*) FROM", tbnm))[[1]]
  # get all column names of the table
  ls_colnm <- c(dbGetQuery(con, paste("SELECT COLUMN_NAME FROM INFORMATION_SCHEMA.COLUMNS WHERE TABLE_NAME='",tbnm,"'",sep=""))$COLUMN_NAME)
  # set up variables for iteration
  colnumber <- length(ls_colnm)
  mincolnbr <- 1
  maxcolnbr <- colnumber
  sp <- mincolnbr
  ep <- maxcolnbr
  # initiate an empty dataframe
  df_total <- setNames(data.frame(matrix(nrow=rownumber, ncol=1)),c("x"))
  
  while (dim(df_total)[2] < maxcolnbr) {
    #print(c(sp,ep))
    df_test <- testpull(con, tbnm, rownumber, ls_colnm, sp, ep)
    #print(typeof(df_test))
    if (typeof(df_test) == "list") {
      # save successfully pulled data
      if (c(colnames(df_total))[1] == "x") {
        df_total <- subset(cbind(df_total, df_test),select=-c(x))
      }
      else {
        df_total <- cbind(df_total, df_test)
      }
      if (dim(df_total)[2] < maxcolnbr) {
        # adjust both start and end points when testpull succeeded
        sp <- ep + 1
        ep <- maxcolnbr
        #print(c(sp, ep))
      }
    }
    else {
      dbDisconnect(con)
      con <- dbConnect(odbc::odbc(), "HistoTrac", timeout = 10)
      # adjust end point when testpull failed
      ep <- floor(sp-1+((ep-(sp-1))/2))
      #print(c(sp, ep))
    }
  }
  return(df_total)
}

(dtTest <- pull(con, "Test") %>% as_tibble)
dtTest %>% write_csv("Test.csv.gz")
```
#Now a copy of the necessary HistoTrac tables are saved as csv files. The next section loads these back into R, and manipulates the data to find families. You can start here if you have the tables saved, and don't need to interface with HistoTrac.
```{r}
#This pulls in the Patient, PatientCase, RelatedPatient, Sample and Test tables from HistoTrac.

(total_patient_table <- read_csv("Patient.csv.gz",
                                 #the arguments below tell R how to interpret the values in several rows, such as typing, which should be treated as characters, not numbers.
                                col_types = cols(
                                  mC1EqCd = col_character(),
                                  drw2cd = col_character(),
                                  drw1cd = col_character(),
                                  A2Cd = col_character(),
                                  mDPB12EqCd = col_character(),
                                  dq2cd = col_character(),
                                  mDQA11EqCd = col_character(),
                                  mDQA12EqCd = col_character(),
                                  mA2EqCd = col_character(),
                                  LowRiskAntibodyLastUpdateDt = col_date(),
                                  mDPB11EqCd = col_character(),
                                  mDQB12EqCd = col_character(),
                                  mB2EqCd = col_character(),
                                  mDRB42EqCd = col_character(),
                                  mB2EqCd = col_character(),
                                  mDRB41EqCd = col_character(),
                                  mDPA12EqCd = col_character(),
                                  mDPA11EqCd = col_character(),
                                  MICA1Cd = col_character(),
                                  MICA2Cd = col_character()
                                )
                                )
)
(PatientCase <- read_csv("PatientCase.csv.gz"))
(RelatedPatient <- read_csv("RelatedPatient.csv.gz"))
(Sample <- read_csv("Sample.csv.gz"))
(Test <- read_csv("Test.csv.gz"))
```

```{r}
#Patients and donors are kept in the same "Patient" table. This separates them to different tables, which makes the subsequent steps easier. Patient/donor is in column "PatientTypeCd"
(patients <- filter(total_patient_table, PatientTypeCd == "Patient"))
(donors <- filter(total_patient_table, PatientTypeCd == "Donor"))
```

```{r}
#This simplifies the tables by selecting certain columns in the patients and donors tables and putting them in new tables called "simple_patients" and "simple_donors"
(simple_patients <- (select(patients, PatientId, HospitalID, firstnm, lastnm, DOB, categoryCd, GenderCd, RaceCd)))
(simple_donors <- (select(donors, PatientId, HospitalID, firstnm, lastnm, DOB, categoryCd, GenderCd, RaceCd)))
```

```{r}
#This links patients with the PatientCase table, then the RelatedPatient table, then back to the total_patient_table. This will add new rows whenever a patient has a donor.  
(linked_cases <- simple_patients %>%
    left_join(PatientCase, by = "PatientId") %>%
    left_join(RelatedPatient, by = "CaseId") %>%
    left_join(total_patient_table, by= c("RelatedPatientID" = "PatientId"))
    )

#This gives you a list of all of the different types of donors in HistoTrac. This is important for the next steps, where you will search for families.
count(linked_cases, RelationshipTypeCd)
```

```{r}
#Using the table above to determine all of the different types of donors we had, I created a new table that only kept the types of patient/donors that would make a family. This will have to be changed, based on what you have in your HistoTrac database.

#This filters to keep only families, as only the relationships below will make families.
(linked_families <-
    linked_cases %>%
    filter(RelationshipTypeCd == "DAUGHTER"|
           RelationshipTypeCd == "Daughter"|
           RelationshipTypeCd == "FATHER" |
           RelationshipTypeCd == "Father"|
           RelationshipTypeCd == "HUSBAND" |
           RelationshipTypeCd == "Husband"|
           RelationshipTypeCd == "MOTHER" |
           RelationshipTypeCd == "Mother"|
           RelationshipTypeCd == "PT" |
           RelationshipTypeCd == "PT."|
           RelationshipTypeCd == "Sibling" |
           RelationshipTypeCd == "SIBLING"|
           RelationshipTypeCd == "SON" |
           RelationshipTypeCd == "Son"|
           RelationshipTypeCd == "WIFE" |
           RelationshipTypeCd == "Wife"|
           RelationshipTypeCd == "Brother" |
           RelationshipTypeCd == "BROTHER"|
           RelationshipTypeCd == "Fiance" |
           RelationshipTypeCd == "FIANCE"|
           RelationshipTypeCd == "SISTER"|
           RelationshipTypeCd == "Sister")
  
 #This simplifies the table.
  %>% select(PatientId:lastnm.x, RaceCd.x, PatientID:RelationshipTypeCd, firstnm.y, lastnm.y, mA1Cd:mDPB12cd)
  
#Our HistoTrac database sometimes had multiple types of donor relationships, which meant we sometimes had patients listed as "PT" and "PT." The code below changes all "PT." to "PT" to make sure all families are included. This may not be required for your database.
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "PT.", "PT"))

#This removes duplicate patient listings
  %>% distinct(PatientID, RelatedPatientID, .keep_all = TRUE)
  
#This changes RelationshipTypeCd to simplify relationships.The idea here is to change all donors to be either "Child, Spouse, Parent, or Sibling," which makes searching in subsequent steps much easier.
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "DAUGHTER", "Child")) 
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "Daughter", "Child"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "FATHER", "Parent"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "Father", "Parent"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "HUSBAND", "Spouse"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "Husband", "Spouse"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "MOTHER", "Parent"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "Mother", "Parent"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "SON", "Child"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "Son", "Child"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "WIFE", "Spouse"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "Wife", "Spouse"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "Brother", "Sibling"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "BROTHER", "Sibling"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "Fiance", "Spouse"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "FIANCE", "Spouse"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "SISTER", "Sibling"))
  %>% mutate(RelationshipTypeCd = replace(RelationshipTypeCd, RelationshipTypeCd == "Sister", "Sibling"))
)
```


#The next set of code links the families table to the test table so we can determine which patients and donors have been tested by NGS. Families are then filtered for those that meet the requirements of the IHIW project.

```{r}
#This links families with the Sample table, then the Test table. 
(families_linked_to_test <- 
   (linked_families %>%
   left_join(Sample, c("RelatedPatientID" = "PatientId"))
   %>% select(PatientId:SampleID, SampleNbr)
   %>% left_join(Test, c("SampleID" = "SampleId"))
   
#Filter for NGS tested samples. You may have a different way to determine who was tested by NGS.
   %>% filter(TestTypeCd == "HR NGS")
   %>% distinct(RelatedPatientID, .keep_all = TRUE)
#Rename columns to make final table more user-friendly
   %>% rename(Patient.HospitalID = HospitalID.x)
   %>% rename(Patient.firstnm = firstnm.x)
   %>% rename(Patient.lastnm = lastnm.x)
   %>% rename(Patient.Race = RaceCd.x)
   %>% rename(Case.firstnm = firstnm.y)
   %>% rename(Case.lastnm = lastnm.y)
#Rearrange columns
  %>% select(PatientID, Patient.HospitalID:Patient.Race, RelatedPatientID:Case.lastnm, SampleID, SampleNbr, mA1Cd.x:mDPB12cd)
   )
 )

#This is where the real magic happens. I essentially realized that there are four types of families that will satisfy the requirements for the IHIW project: "Patient, Spouse and Child," "Patient and 2 children," "Patient and 2 Parents," or "Patient, Parent, and Sibling." The code below groups the families, and then keeps only those that match one of the four criteria. 

#Find families when patient is parent, including spouse
(families_NGSa <- families_linked_to_test %>%
   group_by(PatientID) %>%
   filter(all(c("PT","Spouse", "Child") %in% RelationshipTypeCd))
              )
#Find families when patient is parent, no spouse
(families_NGSb <- families_linked_to_test %>%
    filter(RelationshipTypeCd == "PT"|
            RelationshipTypeCd == "Child")
   %>% group_by(PatientID) 
   %>% filter(n() >=3)
              )
#Find families when patient is child, no siblings
(families_NGSc <- families_linked_to_test %>%
    filter(RelationshipTypeCd == "PT"|
            RelationshipTypeCd == "Parent")
  %>% group_by(PatientID) 
  %>% filter(n() >=3)
              )
#Find families when patient is child, including sibling
(families_NGSd <- families_linked_to_test %>%
   group_by(PatientID) %>%
   filter(all(c("PT", "Parent", "Sibling") %in% RelationshipTypeCd))
              )
#Join all the tables into one
(families_NGS_with_duplicates <- bind_rows(families_NGSa, families_NGSb, families_NGSc, families_NGSd))

#Remove all the duplicate entries, arrange by patient so families show up in groups
(families_NGS <- families_NGS_with_duplicates 
  %>% distinct(RelatedPatientID, .keep_all = TRUE)
  %>% arrange(PatientID))
 
#Save table to files, both csv and Excel files
write_excel_csv(families_NGS, "families_NGS.txt")
write.xlsx(families_NGS, file = "families_NGS.xlsx")

#There you go, you have a list of all families that meet the IHIW requirements, and that have been tested by NGS. You can open the files in Excel.
```

```{r}
#This will count all of the families found (that number will be the number of rows; the "n" value is the number of members in each family). 
count(families_NGS, PatientID)

#Count the types of families from above

#count families when patient is parent, including spouse
families_NGS %>%
   group_by(PatientID) %>%
   filter(all(c("PT","Spouse", "Child") %in% RelationshipTypeCd)) %>%
    count(PatientID)
#Count families when patient is parent, no spouse
families_NGS %>%
    filter(RelationshipTypeCd == "PT"|
            RelationshipTypeCd == "Child") %>% 
   group_by(PatientID) %>% 
   filter(n() >=3) %>%
   count(PatientID)
#Count families when patient is child, no siblings
families_NGS %>%
    filter(RelationshipTypeCd == "PT"|
            RelationshipTypeCd == "Parent") %>% 
  group_by(PatientID) %>%
  filter(n() >=3) %>%
  count(PatientID)
#Count families when patient is child, including sibling
families_NGS %>%
   group_by(PatientID) %>%
   filter(all(c("PT", "Parent", "Sibling") %in% RelationshipTypeCd)) %>%
    count(PatientID)
```
              